---
title: "R Notebook"
output: html_notebook
---
# Required Packages
```{r}
require(dplyr)
require(ggplot2)
require(gplots)
require(MASS)
require(qualityTools)
require(fitdistrplus)
require(DescTools)
require(corrplot)
require(reshape2)
require(heatmaply)
```

# Setup
```{r warning=FALSE}
setwd("C:/Users/isaac/CAPSTONE-STAT-482/Datasets")
paths = read.csv('paths.csv')
players = read.csv('players.csv')
setwd("C:/Users/isaac/CAPSTONE-STAT-482/R-STAT-482-Capstone")
```

# Trim Duplicate Columns and Merge Data
```{r}
# remove repetitive and unnecessary columns in paths.csv
paths = select(paths, -valid_seasons, -link)
# remove duplicate observations
paths = distinct(paths) 

# remove repetitive and unnecessary columns in players.csv
players = select(players, -college, -link, -rank, -draft_year, -draft_rd, -draft_pk, -recruit_year, -bbrID, -pick_overall, -pick_round, -pick_number)
# remove duplicate observations
players = distinct(players)
```

# Merge two data frames into one
```{r}
data = inner_join(paths, players, by = "name")
```

# Change data types 
```{r}
# data$college = as.factor(data$college)
data$smallMult = as.factor(data$smallMult)
data$smallMult = as.factor(data$smallMult)
data$draft_year = as.factor(data$draft_year)
data$draft_rd = as.factor(data$draft_rd)
data$recruit_year = as.factor(data$recruit_year)
data$top = as.factor(data$top)
data$highSchool = as.factor(data$highSchool)
data$coll = as.factor(data$coll)
data$draft = as.factor(data$draft)
data$rookie = as.factor(data$rookie)
data$bad = as.factor(data$bad)
data$good = as.factor(data$good)
data$great = as.factor(data$great)
data$allstar = as.factor(data$allstar)
data$highest = as.factor(data$highest)
```
# Re order factors for "highest" and rename various columns
```{r}
data$highest <- factor(data$highest, levels = c("highSchool", "college", "draft", "rookie", "bad", "good", "great", "allstar"))
# data = rename(data, hsrank = rank)
# data = rename(data, avgNBArank = avgRank)
# data = rename(data, top100 = top)
str(data)
```

# 5 Number Summary (also check number of NAs)
```{r}
summary(data)
```
# Plot the Highest Level Reached by Draft Round
```{r}
college_funnel = ggplot(data = subset(data, !is.na(draft_rd)), aes(x=highest)) + geom_bar(aes(fill=draft_rd)) + ggtitle("Bar Chart Highest Level Reached by Draft Round") + theme(plot.title = element_text(hjust = 0.5)) + xlab("highest level reached")#remove na rows for draft_rd
plot(college_funnel)
```
# Distribution of avgRank
```{r}
avgRank_distr = ggplot(data, aes(x=avgRank)) + geom_histogram(binwidth = 5, fill="dark green")  + ggtitle("Histogram of Average NBA Rank") + theme(plot.title = element_text(hjust = 0.5))
plot(avgRank_distr)
```
# Kernel Density Estimate of avgRank
```{r}
plot(density(data$avgRank, na.rm=TRUE, kernel = "gaussian", bw="nrd0"), main = "Kernel Density Estimate of avgRank with Gaussian Kernel", xlab = "Average NBA Rank")
```

# Distribution of total_seasons
```{r}
total_seasons_distr = ggplot(data, aes(x=total_seasons)) + geom_histogram(binwidth = 1,position="dodge", fill="blue")  + ggtitle("Histogram of Total Seasons Played") + theme(plot.title = element_text(hjust = 0.5))+xlab("Total NBA seasons played")
plot(total_seasons_distr) 
```
# Kernel Density Estimate of total_seasons
```{r}
plot(density(data$total_seasons, na.rm=TRUE, kernel = "gaussian", bw="nrd0"), main = "Kernel Density Estimate of Total Seasons (Gaussian Kernel)", xlab = "Total NBA seasons played")
```
# Fit Poisson Curve to Model Using MLE 
```{r}
total_season_rmna = as.vector(na.omit(as.vector(data$total_seasons)))
poisson_fit = fitdistr(total_season_rmna, "poisson")
poisson_fit
```
# Fit qqplot for poisson model to check goodness of fit
```{r}
# qqPlot(total_season_rmna, "Poisson", main = "QQPlot for Poisson Distirbution Model")
plot(fitdist(total_season_rmna,"pois"))
```
# Fit qqplot
```{r}
PlotQQ(total_season_rmna, function(p) qpois(p, 5.05201699), main = "QQPlot for Total Seasons: Poisson Distribution")
```




# Plot the Distribution of players across draft year
```{r}
draft_year_distr = ggplot(data, aes(x=draft_year)) + geom_bar(fill="light blue") + ylim(0,100)
plot(draft_year_distr)
```
# Plot the Distribution of highest level reached across time
```{r}
highest_time_distr = ggplot(data = subset(data, !is.na(draft_year)), aes(x=draft_year)) + geom_bar(aes(fill=highest)) + ylim(0,100) + ggtitle("Bar Chart of Players per Draft Class by Highest Level Reached") + theme(plot.title = element_text(hjust = 0.5))+xlab("Draft Year")
plot(highest_time_distr)
```
# Plot the Highest Level Reached for draft pick
```{r}
draft_pick_level = ggplot(data, aes(y = stat(count) / sum(count),x=draft_pk)) + geom_histogram(color="black",aes(fill=highest),binwidth=5, size=0.05) + ggtitle("Histogram of Draft Pick by Highest Level Reached") + theme(plot.title = element_text(hjust = 0.5))+xlab("Draft Pick Number")
plot(draft_pick_level) + ylab("Relative Frequency")
```
# Scatterplot of player's mean win shares per 48 minutes versus high school rank
```{r}
ggplot(data, aes(x=hsrank, y=nba_mean_ws48)) + geom_point()
```
# Boxplot of high school rank by highest level reached
```{r}
ggplot(data, aes(x=highest, y=hsrank)) + geom_boxplot(outlier.shape = NA) + geom_jitter(width = 0.15, alpha = 0.15) + ggtitle("Box Plots of High School Rank by Highest Level Reached") + theme(plot.title = element_text(hjust = 0.5))+xlab("Highest Level Reached") + ylab("High School Rank")
```

# Scatterplot of total seasons played versus high school rank
```{r}
ggplot(data, aes(x=hsrank, y=total_seasons)) + geom_point() + ggtitle("Scatterplot of Total Seasons Played vs High School Rank") + theme(plot.title = element_text(hjust = 0.5))+xlab("High School Rank") +ylab("Total NBA Seasons Played")
```
# Scatterplot of draft pick versus high school rank
```{r}
ggplot(data, aes(x=hsrank, y=draft_pk)) + geom_point()+ ggtitle("Scatterplot of Draft Pick vs High School Rank") + theme(plot.title = element_text(hjust = 0.5))+xlab("High School Rank") +ylab("Draft Pick Number")
```

# Scatterplot of avgRank versus high school rank
```{r}
ggplot(data, aes(x=hsrank, y=avgNBArank)) + geom_point()+ ggtitle("Scatterplot of Average NBA Rank vs High School Rank") + theme(plot.title = element_text(hjust = 0.5))+xlab("High School Rank") +ylab("Average NBA Rank")
```
# Scatterplot of avgRank versus draft pick number
```{r}
ggplot(data, aes(x=draft_pk, y=avgRank)) + geom_point()
```
# Scatterplot of total seasons versus draft pick number
```{r}
ggplot(data, aes(x=draft_pk, y=total_seasons)) + geom_point()
```
# Correlation matrix for numeric data
```{r}
only_numeric_data = dplyr::select(data, hsrank, draft_pk, avgNBArank, nba_mean_ws48:valid_seasons)
cormat = round(cor(only_numeric_data, method = c("pearson"), use="pairwise.complete.obs"),2)
cormat
corrplot(cormat)

```
# try another correlation heatmap
```{r}
# Get lower triangle of the correlation matrix
  get_lower_tri<-function(cormat){
    cormat[upper.tri(cormat)] <- NA
    return(cormat)
  }
  # Get upper triangle of the correlation matrix
  get_upper_tri <- function(cormat){
    cormat[lower.tri(cormat)]<- NA
    return(cormat)
  }

melted_cormat <- melt(only_numeric_data.cor)
ggplot(data = melted_cormat, aes(x=Var1, y=Var2, fill=value)) + 
  geom_tile()

upper_tri <- get_upper_tri(cormat)

# Melt the correlation matrix
library(reshape2)
melted_cormat <- melt(upper_tri, na.rm = TRUE)
# Heatmap
library(ggplot2)
ggplot(data = melted_cormat, aes(Var2, Var1, fill = value))+
 geom_tile(color = "white")+
 scale_fill_gradient2(low = "blue", high = "red", mid = "white", 
   midpoint = 0, limit = c(-1,1), space = "Lab", 
   name="Pearson\nCorrelation") +
  theme_minimal()+ 
 theme(axis.text.x = element_text(angle = 45, vjust = 1, 
    size = 12, hjust = 1))+
 coord_fixed()


# reorder correlation matrix
reorder_cormat <- function(cormat){
# Use correlation between variables as distance
dd <- as.dist((1-cormat)/2)
hc <- hclust(dd)
cormat <-cormat[hc$order, hc$order]
}

# Reorder the correlation matrix
cormat <- reorder_cormat(cormat)
upper_tri <- get_upper_tri(cormat)
# Melt the correlation matrix
melted_cormat <- melt(upper_tri, na.rm = TRUE)
# Create a ggheatmap
ggheatmap <- ggplot(melted_cormat, aes(Var2, Var1, fill = value))+
 geom_tile(color = "white")+
 scale_fill_gradient2(low = "blue", high = "red", mid = "white", 
   midpoint = 0, limit = c(-1,1), space = "Lab", 
    name="Pearson\nCorrelation") +
  theme_minimal()+ # minimal theme
 theme(axis.text.x = element_text(angle = 45, vjust = 1, 
    size = 12, hjust = 1))+
 coord_fixed()
# Print the heatmap
print(ggheatmap)

## Add correlation coefficeints on the heat map

ggheatmap + 
geom_text(aes(Var2, Var1, label = value), color = "black", size = 4) +
theme(
  axis.title.x = element_blank(),
  axis.title.y = element_blank(),
  panel.grid.major = element_blank(),
  panel.border = element_blank(),
  panel.background = element_blank(),
  axis.ticks = element_blank(),
  legend.justification = c(1, 0),
  legend.position = c(0.6, 0.7),
  legend.direction = "horizontal")+
  guides(fill = guide_colorbar(barwidth = 7, barheight = 1,
                title.position = "top", title.hjust = 0.5))
```



# simple linear model for total_seasons
```{r}
fit1 = lm(total_seasons ~ log(draft_pk) + log(rank), data = data)
summary(fit1)
plot(fit1)
```
# simple linear model for avgRank
```{r}
fit2 = lm(avgRank ~ log(draft_pk) + log(rank), data = data)
summary(fit2)
plot(fit2)
```
# simple linear model for draft_pk
```{r}
fit3 = lm(draft_pk ~ log(rank), data = data)
summary(fit3)
plot(fit3)
```
# simple linear model for mean win shares per 48 minutes
```{r}
fit4 = lm(nba_mean_ws48 ~ rank + draft_pk, data = data)
summary(fit4)
plot(fit4)
```
